Group Theory in Physics (Course Note)

### 2016-11-12 Lecture note - Group Theory in Physics/sections/20161031.tex
Finished another proof for "conjugacy class"="irrep" in group representation theory.

\begin{thm}[Orthogonality Theorem for Representations]
    For finite group $G$, let $D^i(G)$ and $D^j(G)$ be its two
    irreducible representation. Then, as a vector in group algebra,
    they have the following orthogonal
    relationship:
    \begin{equation}
        \sum_{h\in G} D^{i}_{\mu\rho}(h^{-1})D^j_{\nu \lambda}(h) =
        \frac{N}{d_j} \delta_{ij} \delta_{\mu\nu}\delta_{\rho\lambda}
    \end{equation}
    $N$ is the order of the group, and $d_j$ is the
    dimension of representation $D^j(G)$. If in addition, the two
    representations are unitary, we have
    \begin{equation}
        \sum_{h\in G} D^{i*}_{\mu\rho}(h)D^j_{\nu \lambda}(h) =
        \frac{N}{d_j} \delta_{ij} \delta_{\mu\nu}\delta_{\rho\lambda}
    \end{equation}
\end{thm}
\begin{proof}
    \textbf{Note}: The following proof is clumsy and only applies for
    the unitary case. For a good proof, please refer to page 44 of
    \cite{Ludeling}.

@@

        Now, consider the class vectors $\mathcal{K}_a$ in certain
        irreducible representation $\mu$:
        \begin{equation}
            D^a_{(\mu)} := \sum_{g\in \mathcal{K}_a} D_{(\mu)}(g)
        \end{equation}
        It commutes with all $D_{(\mu)}(h)$, $h\in G$, since
        $$D_{(\mu)}(h) \sum_{g\in \mathcal{K}_a} D_{(\mu)}(g)
        D_{(\mu)}(h^{-1})= \sum_{g\in \mathcal{K}_a} D_{(\mu)}(g)$$
        Then by Schur's lemma, $D^a_{(\mu)}= \lambda^a_\mu
        \id$.
        Take the trace of $D^a_{(\mu)}$, one will easily find:
        \begin{equation}
            \label{eq:20161031-lambda-is}
            \lambda^a_\mu = \frac{n_a\chi^a_{\mu}}{\chi^1_\mu}
        \end{equation}
        Meanwhile, by the algebra like relation
        \ref{eq:20161031-algebra-like-relation}, one finds:
        \begin{equation}
            \lambda^a_\mu \lambda^b_\mu = \sum_c c_{abc} \lambda^c_\mu
        \end{equation}
        combine this with the expression for $\lambda$:
        \ref{eq:20161031-lambda-is}, one has:
        \begin{equation}
            n_a n_b \chi^a_\mu \chi^b_\mu = \chi^1_\mu\sum_c
            c_{abc}n_c\chi^c_\mu
        \end{equation}
        Now, one can see this pattern and I would not expatiate on
        it. The only remaining elements are equation
        \ref{eq:20161031-x1xa-Ndelta} and the fact that
        \begin{equation}
            c_{abe} = \delta_{b,a^{-1}} n_a
        \end{equation}
        where $e$ is the unit element in group $G$, and $a^{-1}$ is
        the inverse class of $a$. This is like saying that only the
        product of $a$ with its inverse class can have unit elements.
        With these one can easily calculate and get
        \begin{equation}
            \sum_{\mu \in \text{I.R.}} \chi^{a*}_\mu\chi^b_\mu = 
            \frac{|G|}{n_a} \delta^{ab}
        \end{equation}
        So we have:
        \begin{equation}
            \sqrt{\frac{n_a}{|G|}}(\chi^a_1,\chi^a_2,\cdots,\chi^a_r)
        \end{equation}
        a series of $k$ orthonomal $r$-dimensional vectors. Hence
        \begin{equation}
            k\leq r
        \end{equation}
        Then, together with previous result, we have:
        \begin{equation}
            r = k
        \end{equation}
    \end{proof}
% 1. 自共轭 iff 特征 实数
% 2. 非自共轭 至少一个群元 复数 -》 成对出现
% 3. 实表示 是 自共轭
% 4. 特征 实数 不一定 实表示 (so 有 自共轭 但非 实的表示)
% 5. thm 5 on pp. 68.

>>>>

\begin{thm}[Orthogonality Theorem for Representations]
    For finite group $G$, let $D^i(G)$ and $D^j(G)$ be its two
    irreducible representation. Then, as a vector in group algebra,
    they have the following orthogonal
    relationship:
    \begin{equation}
        \sum_{h\in G} D^{i}_{\mu\rho}(h^{-1})D^j_{\nu \lambda}(h) =
        \frac{N}{d_j} \delta_{ij} \delta_{\mu\nu}\delta_{\rho\lambda}
        \label{eq:20161031-ortho-dd-1}
    \end{equation}
    $N$ is the order of the group, and $d_j$ is the
    dimension of representation $D^j(G)$. If in addition, the two
    representations are unitary, we have
    \begin{equation}
        \sum_{h\in G} D^{i*}_{\mu\rho}(h)D^j_{\nu \lambda}(h) =
        \frac{N}{d_j} \delta_{ij} \delta_{\mu\nu}\delta_{\rho\lambda}
        \label{eq:20161031-ortho-dd-2}
    \end{equation}
\end{thm}
\begin{proof}
    \textbf{Note}: The following proof is clumsy and only applies for
    the unitary case. For a good proof, please refer to page 44 of
    \cite{Ludeling}.

@@

        Now, consider the class vectors $\mathcal{K}_a$ in certain
        irreducible representation $\mu$:
        \begin{equation}
            D^a_{(\mu)} := \sum_{g\in \mathcal{K}_a} D_{(\mu)}(g)
        \end{equation}
        It commutes with all $D_{(\mu)}(h)$, $h\in G$, since
        $$D_{(\mu)}(h) \sum_{g\in \mathcal{K}_a} D_{(\mu)}(g)
        D_{(\mu)}(h^{-1})= \sum_{g\in \mathcal{K}_a} D_{(\mu)}(g)$$
        Then by Schur's lemma, $D^a_{(\mu)}= \lambda^a_\mu
        \id$.
        Take the trace of $D^a_{(\mu)}$, one will easily find:
        \begin{equation}
            \label{eq:20161031-lambda-is}
            \lambda^a_\mu = \frac{n_a\chi^a_{\mu}}{\chi^1_\mu}
        \end{equation}
        Meanwhile, by the algebra like relation
        \ref{eq:20161031-algebra-like-relation}, one finds:
        \begin{equation}
            \lambda^a_\mu \lambda^b_\mu = \sum_c c_{abc} \lambda^c_\mu
        \end{equation}
        combine this with the expression for $\lambda$:
        \ref{eq:20161031-lambda-is}, one has:
        \begin{equation}
            n_a n_b \chi^a_\mu \chi^b_\mu = \chi^1_\mu\sum_c
            c_{abc}n_c\chi^c_\mu
        \end{equation}
        Now, one can see this pattern and I would not expatiate on
        it. The only remaining elements are equation
        \ref{eq:20161031-x1xa-Ndelta} and the fact that
        \begin{equation}
            c_{abe} = \delta_{b,a^{-1}} n_a
        \end{equation}
        where $e$ is the unit element in group $G$, and $a^{-1}$ is
        the inverse class of $a$. This is like saying that only the
        product of $a$ with its inverse class can have unit elements.
        With these one can easily calculate and get
        \begin{equation}
            \sum_{\mu \in \text{I.R.}} \chi^{a*}_\mu\chi^b_\mu = 
            \frac{|G|}{n_a} \delta^{ab}
            \label{eq:20161031-ortho-character}
        \end{equation}
        So we have:
        \begin{equation}
            \sqrt{\frac{n_a}{|G|}}(\chi^a_1,\chi^a_2,\cdots,\chi^a_r)
        \end{equation}
        a series of $k$ orthonomal $r$-dimensional vectors. Hence
        \begin{equation}
            k\leq r
        \end{equation}
        Then, together with previous result, we have:
        \begin{equation}
            r = k
        \end{equation}
    \end{proof}
    Another proof provided by the book \cite{book}.
    \begin{defi}[Group function]
    \nomenclature{Group function}{\nomrefpage.}
        A $\mathbb{C}$-linear function $F:\mathbb{C}[G]\to
        \mathbb{C}$ is called a group function.
    \end{defi}
    He first shows that:
    \begin{thm}
        The matrix elements of unitary irreducible representations of
        finite groups may form an complete orthogonal basis of group
        algebra. Any group function can be expanded in terms of them.
    \end{thm}
    \begin{proof}
        Let $D^j_{\mu\nu}(g)$, where $g\in G$, $j$ labels irreducible
        representations, $\mu\nu$ labels matrix element, be one such
        element. Then we have, for any function $F:G\to \mathbb{C}$,
        \begin{align}
            F(h) &= \sum_{j,\mu\nu} C^j_{\mu\nu} D^j_{\mu\nu}(h) \\
            C^j_{\mu\nu} &= \frac{d_j}{|G|}
            \sum_{h\in G} D^j_{\mu\nu}(h)^* F(h)
            \label{eq:20161031-cjmv}
        \end{align}
        This can be easily get using the orthonomal relationship.
        Also notes that we have $\sum_j d^2_j = |G| =
        \operatorname{Dim}(\mathbb{C}[G])$, confirming that the
        dimension of basis agree with the dimension of group algebra.
    \end{proof}
    \begin{defi}[Class function]
    \nomenclature{Class function}{\nomrefpage.}
        A class function is a group function $f$, whose satisfies:
        \begin{equation}
            f(s g s^{-1}) = f(g)
        \end{equation}
        for any $s,g\in G$.
    \end{defi}
    Similarly, we have:
    \begin{thm}
        The character $\chi^j(g)$ ,where $j$ labels irreducible
        representation, $g\in G$, of unitary irreducible representations of
        finite groups may form an complete orthogonal basis of class
        functions.
    \end{thm}
    \begin{proof}
        This is easy if one realize that, since $f(sgs^{-1}=f(g)$ for
        any $s,g\in G$, one has
        \begin{equation}
            f(g) = \frac{1}{|G|} \sum_{s\in G} f(s g s^{-1})
        \end{equation}
        The rest is to use result mentioned in proving the group
        function basis above. One finally get:
        \begin{equation}
            f(g) = \sum_j \left(\frac{1}{d_j}\sum_\mu
            C^j_{\mu\mu}\right) \chi^j(g)
        \end{equation}
        where $C^j_{\mu\mu}$ is defined in equation
        \ref{eq:20161031-cjmv}.
        Or we can write it in another way:
        \begin{align}
            F(g) &= \sum_j C_j \chi^j(g) \\
            C_j &= \frac{1}{|G|} \sum_{h\in G}\chi^j(h)^* F(h)
        \end{align}
        the formula for $C_j$ is also easily found by simply
        calculating $\sum_{h\in G}\chi^j(h)^* F(h)$.
    \end{proof}
    Note that the above result is enough to show that the number of
    irreducible representations is the same as the dimension of
    class functions (consider it as a vector space).

    The book \cite{book} also rephrase it in page 63, as:
    \begin{quote}
        Define two matrices $U_{m,n}$ and $V_{m,n}$, whose elements
        are defined by:
    \begin{align}
        U_{g,j\mu\nu} &:= \sqrt{\frac{d_j}{|G|}} D^j_{\mu\nu}(g) \\
        V_{a,j} &:= \sqrt{\frac{n_a}{|G|}} \chi^j_a
    \end{align}
    The two matrices are unitary. This is true by equation
    \ref{eq:20161031-ortho-dd-2} and equation
    \ref{eq:20161031-ortho-character}. Note here $g\in G$, $j$ labels
    different unitary irreducible representations, $\mu,\nu$ labels
    the matrix element in that representation, $a$ labels conjugacy
    classes. The labels $j\mu\nu$ together form the second index of
    matrix $U_{m,n}$. For convenience, I type all the orthonomal
    relationship together:
    \begin{align}
        &\text{Equation \ref{eq:20161031-ortho-dd-2}:} \nonumber \\
        & \sum_{h\in G} D^{i*}_{\mu\rho}(h)D^j_{\nu \lambda}(h) =
        \frac{N}{d_j} \delta_{ij} \delta_{\mu\nu}\delta_{\rho\lambda}
        \\
        &\text{Equation \ref{eq:20161031-ortho-character}:} \nonumber \\
        & \sum_{\mu \in \text{I.R.}} \chi^{a*}_\mu\chi^b_\mu = 
            \frac{|G|}{n_a} \delta^{ab} 
    \end{align}

    \end{quote}

% 1. 自共轭 iff 特征 实数
% 2. 非自共轭 至少一个群元 复数 -》 成对出现
% 3. 实表示 是 自共轭
% 4. 特征 实数 不一定 实表示 (so 有 自共轭 但非 实的表示)
% 5. thm 5 on pp. 68.
